---
title: "Correlation-based Network Pipelines"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Correlation-based Network Pipelines}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 5,
  dpi       = 72
)
```

## Overview

This vignette walks through the correlation-network family of builders in
`ggNetView`. These functions turn one or more numeric abundance / expression
tables into a unified, module-annotated graph object that can be plotted
directly with `ggNetView()`.

The three main entry points are:

* `build_graph_from_mat()` — single matrix, correlation network across its
  rows (e.g. OTUs, genes, metabolites).
* `build_graph_from_double_mat()` — two matrices (two feature blocks)
  measured on the same samples; edges represent cross-block correlations.
* `build_graph_from_multi_mat()` — generalizes the double-matrix case to
  three or more feature blocks.

All three return a `tidygraph` object carrying correlation sign, edge
weight, and a deterministic module assignment.

```{r setup}
library(ggNetView)
```

## 1. Single-matrix correlation network

`build_graph_from_mat()` expects a numeric matrix with **features in rows**
and **samples in columns**. The package ships `otu_rare_relative`, a
relative-abundance OTU table. For a fast example we subset to the 60 most
abundant OTUs.

```{r single-mat}
data(otu_rare_relative)

mat <- as.matrix(otu_rare_relative)
row_sums <- rowSums(mat)
mat <- mat[order(row_sums, decreasing = TRUE)[seq_len(60)], ]

graph_obj <- build_graph_from_mat(
  mat           = mat,
  method        = "cor",
  cor.method    = "spearman",
  proc          = "BH",
  r.threshold   = 0.6,
  p.threshold   = 0.05,
  module.method = "Fast_greedy",
  seed          = 1
)

graph_obj
```

### Choosing an inference method

The `method` argument selects how edges are inferred:

| `method`     | Engine                                     | P-values | Typical use |
|--------------|--------------------------------------------|----------|-------------|
| `"cor"`      | `psych::corr.test()`                       | yes      | General-purpose |
| `"Hmisc"`    | `Hmisc::rcorr()`                           | yes      | Large matrices, Pearson / Spearman only |
| `"WGCNA"`    | `WGCNA::corAndPvalue()`                    | yes      | Gene co-expression |
| `"SPARCC"`   | Internal Rcpp SparCC + bootstrap           | yes      | Compositional microbiome data |
| `"SpiecEasi"`| Internal Rcpp SpiecEasi (`mb` / `glasso`)  | no       | Sparse inverse-covariance |

Switch methods by changing a single argument — the rest of the pipeline
(`r.threshold`, module detection, attribute assembly) stays the same.

### Attaching taxonomy

Passing a node annotation table merges metadata onto the vertices. The
first column of the annotation must match the feature names.

```{r single-mat-annot}
data(tax_tab)

annot <- tax_tab[tax_tab$OTUID %in% rownames(mat), ]
graph_obj_annot <- build_graph_from_mat(
  mat             = mat,
  method          = "cor",
  cor.method      = "spearman",
  proc            = "BH",
  r.threshold     = 0.6,
  p.threshold     = 0.05,
  module.method   = "Fast_greedy",
  node_annotation = annot,
  seed            = 1
)
```

Downstream plots can now colour or facet by any taxonomy column
(`Phylum`, `Class`, …).

## 2. Visualising the network

`ggNetView()` is the single plotting entry point. Edge colour follows
correlation sign automatically; node fill can be driven by `Modularity`
or any column attached via `node_annotation`.

```{r plot-single}
ggNetView(
  graph_obj_annot,
  layout    = "fr",
  seed      = 1,
  node_size_range = c(2, 7),
  node_fill   = "Modularity",
  module_label     = FALSE
)
```

Swap layouts by changing the `layout` string — every layout uses the
supplied `seed` so figures are reproducible.

```{r plot-single-circle}
ggNetView(
  graph_obj_annot,
  layout    = "circle",
  seed      = 1,
  node_size_range = c(2, 7),
  node_fill   = "Modularity"
)
```

## 3. Two-block (double-matrix) networks

Use `build_graph_from_double_mat()` when you have two feature blocks
measured on the same samples and want edges that only cross between the
two blocks (e.g. bacteria ↔ fungi, or microbes ↔ metabolites).

Both matrices must share the same sample column names.

```{r double-mat}
data(BASV_tab)
data(FASV_tab)

mat_b <- as.matrix(BASV_tab)
mat_f <- as.matrix(FASV_tab)

double_obj <- build_graph_from_double_mat(
  mat1          = mat_b,
  mat2          = mat_f,
  module.method = "Fast_greedy",
  seed          = 1
)

double_obj
```

The returned graph behaves identically to the single-matrix case, so the
same `ggNetView()` call works:

```{r plot-double}
ggNetView(
  double_obj,
  layout    = "fr",
  seed      = 1,
  node_size_range = c(2, 6),
  node_fill   = "Modularity",
  module_label     = FALSE
)
```

For a bipartite-style rendering, swap `layout = "fr"` for
`"bipartite"` and supply the block assignment via `node_annotation`.

## 4. Multi-block networks

`build_graph_from_multi_mat()` is the natural extension to three or more
blocks. Matrices are passed positionally or via `...`; the function
intersects sample names and stacks features.

```{r multi-mat}
set.seed(1)
nsamp <- 20
mat_a <- matrix(stats::rnorm(10 * nsamp), nrow = 10)
mat_b <- matrix(stats::rnorm(10 * nsamp), nrow = 10)
mat_c <- matrix(stats::rnorm(10 * nsamp), nrow = 10)
rownames(mat_a) <- paste0("A", seq_len(10))
rownames(mat_b) <- paste0("B", seq_len(10))
rownames(mat_c) <- paste0("C", seq_len(10))
colnames(mat_a) <- colnames(mat_b) <- colnames(mat_c) <-
  paste0("sample", seq_len(nsamp))

multi_obj <- build_graph_from_multi_mat(
  mat_a, mat_b, mat_c,
  module.method = "Fast_greedy",
  seed          = 1
)

multi_obj
```

## 5. Inspecting the result

All three builders return a `tidygraph` object, so node- and edge-level
tibbles are directly accessible via `get_graph_nodes()` and
`get_info_from_graph()`:

```{r inspect}
head(get_graph_nodes(graph_obj))
lapply(get_info_from_graph(graph_obj), head, 3)
```

Every node carries:

* `Modularity` — top-k module label (`Others` bucket when many small modules)
* `Degree`, `Strength` — degree and weighted degree
* any column merged in through `node_annotation`

Every edge carries:

* `weight` — absolute correlation
* `correlation` — signed correlation
* `corr_direction` — `"Positive"` / `"Negative"`

## 6. Network topology

`get_network_topology()` computes global network metrics plus a
bootstrap-based robustness summary. It accepts either a pre-built graph
object or a raw matrix — in the latter case the builder is re-run
internally so that the same `r.threshold` / `p.threshold` / `method`
combination is applied.

```{r topology}
topo <- get_network_topology(graph_obj = graph_obj, bootstrap = 20)
names(topo)
head(topo$topology)
```

The return value is a list with at least two elements:

* `topology` — one row of global metrics (number of nodes / edges,
  average degree, clustering coefficient, modularity, …).
* `robustness` — bootstrap summary of how network-level metrics behave
  under random node removal, used to estimate stability.

For sample × feature designs (e.g. microbiome studies with many
samples) use `get_sample_subgraph_topology()` or its `_parallel`
variant to compute the same metrics per-sample.

## 7. Keystone detection with Zi-Pi

`ggnetview_zipi()` classifies every node by its within-module
connectivity (Zi) and among-module participation coefficient (Pi).
The default thresholds (`Zi = 2.5`, `Pi = 0.62`) are those of Guimerà
& Amaral (2005) and split nodes into four roles:

* **Peripherals** — low Zi, low Pi
* **Connectors** — low Zi, high Pi
* **Module hubs** — high Zi, low Pi
* **Network hubs** — high Zi, high Pi

It takes a node table and an adjacency matrix — both of which are
accessible directly from a `ggNetView` graph object:

```{r zipi}
nodes_tbl <- get_graph_nodes(graph_obj)
adj_mat   <- get_graph_adjacency(graph_obj)

zipi <- ggnetview_zipi(
  nodes_bulk     = nodes_tbl,
  z_bulk_mat     = adj_mat,
  modularity_col = "Modularity",
  degree_col     = "Degree"
)

head(zipi$data[, c("name", "within_module_connectivities",
                   "among_module_connectivities", "type")])
```

The `plot` element in the result is a ready-to-render Zi-Pi scatter
plot with the four quadrants shaded:

```{r zipi-plot}
zipi$plot
```

## 8. Where to go next

* **Layout gallery**: pass any `layout = "..."` string accepted by
  `ggNetView()` — there are 30+ deterministic layouts, including
  `"circular_modules_petal"`, `"cross_quadripartite_gephi"`, and
  `"star_concentric"`.
* **Module comparison across networks**: build a named list of graph
  objects and feed it to `get_network_topology(graph_obj_list = ...)`
  to get per-network metrics in one call; use
  `ggnetview_modularity_heatmaps()` to visualise module composition
  side-by-side.
* **RMT thresholding**: `ggNetView_RMT()` picks a data-driven
  correlation threshold using random matrix theory — useful when you
  do not want to hand-tune `r.threshold`.
* **Multi-network rendering**: `ggNetView_multi()` /
  `ggNetView_multi_link()` plot several networks in a shared
  coordinate system with aligned modules.

## Session information

```{r sessioninfo}
sessionInfo()
```
