---
title: "Introduction to ggNetView"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Introduction to ggNetView}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 5,
  dpi       = 72
)
```

## Overview

`ggNetView` provides a reproducible, deterministic framework for analyzing and
visualizing biological, ecological, and microbial association networks. The
package combines:

* a family of `build_graph_from_*()` functions that turn adjacency matrices,
  correlation matrices, edge lists, or `igraph` objects into a unified
  `tidygraph`-backed network object with module annotations;
* topology utilities (`get_network_topology()`, `get_info_from_graph()`,
  `get_graph_nodes()`) for node- and network-level metrics;
* a large collection of deterministic layout generators (dispatched through
  `ggNetView()` via the `layout` argument); and
* a `ggplot2`-compatible plotting function `ggNetView()` for
  publication-quality figures.

This vignette walks through an end-to-end workflow: loading a bundled example,
building a graph object, extracting information, and rendering a plot.

## Load the package

```{r setup}
library(ggNetView)
library(igraph)
```

## 1. Build a graph object

`ggNetView` ships with several example datasets. We use `ppi_example`, a
small protein–protein interaction network with 100 nodes and 200 edges.

```{r build}
data(ppi_example)

ig <- igraph::graph_from_data_frame(
  d        = ppi_example$ppi,
  vertices = ppi_example$annotation,
  directed = FALSE
)

graph_obj <- build_graph_from_igraph(
  igraph        = ig,
  module.method = "Fast_greedy"
)

graph_obj
```

`build_graph_from_igraph()` detects communities (here with the Fast-greedy
algorithm), simplifies any multi-edges or self-loops, and attaches
node-level attributes (`Modularity`, `Degree`, `Strength`) for downstream
plotting.

## 2. Inspect nodes and edges

Use `get_graph_nodes()` for a quick node tibble, or `get_info_from_graph()`
for both nodes and edges as a list.

```{r inspect}
nodes <- get_graph_nodes(graph_obj)
head(nodes)

info <- get_info_from_graph(graph_obj)
lapply(info, head, 3)
```

## 3. Visualize with `ggNetView()`

The `ggNetView()` function is the main entry point for plotting. The `layout`
argument selects one of the deterministic layout generators shipped with the
package. Below we render the network with the Fruchterman–Reingold layout and
color nodes by module.

```{r plot-fr}
ggNetView(
  graph_obj,
  layout    = "fr",
  seed      = 1,
  node_size_range = c(2, 8),
  node_fill   = "Modularity",
  module_label     = FALSE
)
```

Switching layouts only requires changing the `layout` string. Here is the same
graph rendered on a circular layout:

```{r plot-circle}
ggNetView(
  graph_obj,
  layout    = "circle",
  seed      = 1,
  node_size_range = c(2, 8),
  node_fill   = "Modularity"
)
```

Every layout in `ggNetView` uses a controllable `seed` so figures are
bit-for-bit reproducible across runs.

## 4. Next steps

For correlation-network pipelines from raw abundance tables, see
`build_graph_from_mat()`, `build_graph_from_double_mat()`, and the
`get_network_topology()` / `get_sample_subgraph_topology()` families. For
module-level comparisons across samples, see
`ggnetview_modularity_heatmaps()` and `ggnetview_zipi()`.

## Session information

```{r sessioninfo}
sessionInfo()
```
