---
title: "Layout Gallery"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Layout Gallery}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 6,
  fig.height = 6,
  dpi       = 72
)
```

## Overview

`ggNetView` ships 60+ deterministic layouts. Every layout is selected
by passing a single string to `ggNetView(layout = "...")`. The `seed`
argument ensures identical node placement across runs.

This vignette is a visual catalogue organised by family. All examples
use the same graph object so differences come purely from the layout.

```{r setup}
library(ggNetView)
```

## Build a demo network

We reuse the bundled OTU data and build a small, well-connected
network with clear modules.

```{r demo-graph}
data(otu_rare_relative)
data(tax_tab)

mat <- as.matrix(otu_rare_relative)
mat <- mat[order(rowSums(mat), decreasing = TRUE)[seq_len(80)], ]
annot <- tax_tab[tax_tab$OTUID %in% rownames(mat), ]

g <- build_graph_from_mat(
  mat             = mat,
  method          = "cor",
  cor.method      = "spearman",
  proc            = "BH",
  r.threshold     = 0.6,
  p.threshold     = 0.05,
  module.method   = "Fast_greedy",
  node_annotation = annot,
  seed            = 1
)
```

A small helper to keep the gallery code concise:

```{r helper}
show_layout <- function(layout_name, ...) {
  ggNetView(
    g,
    layout    = layout_name,
    seed      = 1,
    node_size_range = c(2, 6),
    node_fill   = "Modularity",
    module_label     = FALSE,
    ...
  ) +
    ggplot2::ggtitle(layout_name)
}
```

---

## 1. Force-directed layouts

Force-directed algorithms simulate physical forces (spring attraction,
node repulsion) to produce organic, readable networks. These are the
most common starting point.

```{r force-fr, fig.cap = "Fruchterman-Reingold (fr)"}
show_layout("fr")
```

```{r force-fr1, fig.cap = "Fruchterman-Reingold variant 1 (fr1)"}
show_layout("fr1")
```

```{r force-fr2, fig.cap = "Fruchterman-Reingold variant 2 (fr2)"}
show_layout("fr2")
```

```{r force-kk, fig.cap = "Kamada-Kawai (kk)"}
show_layout("kk")
```

```{r force-stress, fig.cap = "Stress majorisation (stress)"}
show_layout("stress")
```

```{r force-nicely, fig.cap = "igraph 'nicely' auto-selection (nicely)"}
show_layout("nicely")
```

```{r force-nicely1, fig.cap = "nicely variant (nicely1)"}
show_layout("nicely1")
```

```{r force-lgl, fig.cap = "Large Graph Layout (lgl)"}
show_layout("lgl")
```

```{r force-gephi, fig.cap = "Gephi-style ForceAtlas2 (gephi)"}
show_layout("gephi")
```

## 2. Geometric layouts

Simple, deterministic shapes useful when structure clarity matters more
than spatial clustering.

```{r geom-circle, fig.cap = "Circle (circle)"}
show_layout("circle")
```

```{r geom-circle-outline, fig.cap = "Circle outline (circle_outline)"}
show_layout("circle_outline")
```

```{r geom-grid, fig.cap = "Grid (grid)"}
show_layout("grid")
```

```{r geom-star, fig.cap = "Star (star)"}
show_layout("star")
```

```{r geom-star-concentric, fig.cap = "Star concentric (star_concentric)"}
show_layout("star_concentric")
```

```{r geom-diamond, fig.cap = "Diamond (diamond)"}
show_layout("diamond")
```

```{r geom-square, fig.cap = "Square (square)"}
show_layout("square")
```

```{r geom-square2, fig.cap = "Square variant (square2)"}
show_layout("square2")
```

```{r geom-rectangle, fig.cap = "Rectangle (rectangle)"}
show_layout("rectangle")
```

```{r geom-rectangle-outline, fig.cap = "Rectangle outline (rectangle_outline)"}
show_layout("rectangle_outline")
```

```{r geom-petal, fig.cap = "Petal (petal)"}
show_layout("petal")
```

```{r geom-petal2, fig.cap = "Petal variant (petal2)"}
show_layout("petal2")
```

```{r geom-heart, fig.cap = "Heart (heart_centered)"}
show_layout("heart_centered")
```

```{r geom-randomly, fig.cap = "Random (randomly)"}
show_layout("randomly")
```

## 3. Hierarchical layouts

```{r hier-rightiso, fig.cap = "Right-isometric layers (rightiso_layers)"}
show_layout("rightiso_layers")
```

## 4. Circular module layouts

These layouts arrange each module into a distinct geometric shape on a
circle, making module boundaries visually explicit. Each comes in a
standard version (module sizes proportional to node count) and an
**equal** version (all modules the same size).

Circular module layouts require `layout_module = "adjacent"` for best
results. They work best with networks that have well-defined,
moderately-sized modules.

```{r cm-examples, eval = FALSE}
# Standard (proportional) variants
show_layout("circular_modules_gephi_layout", layout_module = "adjacent")
show_layout("circular_modules_petal_layout", layout_module = "adjacent")
show_layout("circular_modules_petal2_layout", layout_module = "adjacent")
show_layout("circular_modules_diamond_layout", layout_module = "adjacent")
show_layout("circular_modules_star_layout", layout_module = "adjacent")
show_layout("circular_modules_star_concentric_layout", layout_module = "adjacent")
show_layout("circular_modules_square_layout", layout_module = "adjacent")
show_layout("circular_modules_square2_layout", layout_module = "adjacent")
show_layout("circular_modules_grid_layout", layout_module = "adjacent")
show_layout("circular_modules_heart_centered_layout", layout_module = "adjacent")

# Equal-sized module variants
show_layout("circular_modules_equal_gephi_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_petal_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_petal2_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_diamond_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_star_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_star_concentric_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_square_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_square2_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_grid_layout", layout_module = "adjacent")
show_layout("circular_modules_equal_heart_centered_layout", layout_module = "adjacent")
```

## 5. Multipartite layouts

Multipartite layouts separate nodes into two or more spatially distinct
groups. Use them when nodes have a categorical block attribute
(e.g. bacteria vs. fungi, or different experimental groups).

The `layout_module` parameter controls how modules are arranged:

- `"random"` -- modules distributed freely
- `"adjacent"` -- modules positioned close together
- `"order"` -- modules follow the block order (required for multipartite)

These layouts require that the number of modules matches the expected
block count (e.g. bipartite needs exactly 2, tripartite needs 3).

```{r multi-examples, eval = FALSE}
# Bipartite (2-block)
show_layout("bipartite_layout", layout_module = "order")
show_layout("bipartite_gephi_layout", layout_module = "order")

# Tripartite (3-block)
show_layout("tripartite_layout", layout_module = "order")
show_layout("tripartite_gephi_layout", layout_module = "order")
show_layout("tripartite_equal_gephi_layout", layout_module = "order")

# Quadripartite (4-block)
show_layout("quadripartite_gephi_layout", layout_module = "order")
show_layout("quadripartite_equal_gephi_layout", layout_module = "order")
show_layout("cross_quadripartite_gephi_layout", layout_module = "order")
show_layout("cross_quadripartite_equal_gephi_layout", layout_module = "order")

# Pentapartite (5-block)
show_layout("pentapartite_gephi_layout", layout_module = "order")
show_layout("pentapartite_equal_gephi_layout", layout_module = "order")
```

## 6. Consensus module layouts

Consensus layouts align modules from different networks into a shared
coordinate system. Useful for comparing network structures across
conditions.

```{r consensus-examples, eval = FALSE}
show_layout("consensus_module_gephi")
show_layout("consensus_module_equal_gephi")
```

## 7. The `layout_module` parameter

All layouts accept `layout_module` to control how modules are spatially
arranged. Here is the same layout with the three options:

```{r lm-random, fig.cap = "layout_module = 'random'"}
show_layout("gephi", layout_module = "random")
```

```{r lm-adjacent, fig.cap = "layout_module = 'adjacent'"}
show_layout("gephi", layout_module = "adjacent")
```

## Quick reference table

| Family | Layout string | Key feature |
|--------|--------------|-------------|
| **Force-directed** | `fr`, `fr1`, `fr2` | Fruchterman-Reingold variants |
| | `kk` | Kamada-Kawai spring model |
| | `stress` | Stress majorisation |
| | `nicely`, `nicely1` | igraph auto-selection |
| | `lgl` | Large graph layout |
| | `gephi` | ForceAtlas2-style |
| **Geometric** | `circle`, `circle_outline` | Circular |
| | `grid` | Regular grid |
| | `star`, `star_concentric` | Star / concentric rings |
| | `diamond`, `diamond_outline` | Diamond shape |
| | `square`, `square2`, `square_outline` | Square shapes |
| | `rectangle`, `rectangle_outline` | Rectangle |
| | `petal`, `petal2` | Petal / flower |
| | `heart_centered` | Heart shape |
| | `randomly` | Random placement |
| **Hierarchical** | `dendrogram` | Circular dendrogram (directed graphs) |
| | `multirings` | Concentric rings |
| | `rightiso_layers` | Right-isometric layers |
| **Circular modules** | `circular_modules_*_layout` | Module per petal/shape |
| | `circular_modules_equal_*_layout` | Equal-sized variant |
| **Multipartite** | `bipartite_layout`, `bipartite_gephi_layout` | 2-block |
| | `tripartite_*` | 3-block |
| | `quadripartite_*`, `cross_quadripartite_*` | 4-block |
| | `pentapartite_*` | 5-block |
| **Consensus** | `consensus_module_gephi` | Aligned multi-network |
| | `consensus_module_equal_gephi` | Equal-sized variant |

## Tips for choosing a layout

1. **Start with `"fr"` or `"gephi"`** for an overview of community
   structure.
2. **Switch to `"circular_modules_*"`** when you want each module
   clearly separated.
3. **Use multipartite layouts** when nodes have a meaningful block
   assignment (bacteria/fungi, treatment/control).
4. **Use `"dendrogram"`** for directed / hierarchical graphs.
5. **Use `layout_module = "adjacent"`** to pull related modules closer
   together.
6. **All layouts are deterministic** when `seed` is set -- figures are
   reproducible across sessions.

## Session information

```{r sessioninfo}
sessionInfo()
```
