---
title: "Virtual Perturbation Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Virtual Perturbation Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi       = 72
)
```

## Overview

Once a network is built, a common question is *what happens to it under
disturbance* -- if key taxa disappear, if one species is pushed up or
down, or if a whole module is lost. `ggNetView` groups these "virtual
perturbation" analyses into three functions that all take a `tbl_graph`
from any `build_graph_from_*()` constructor:

| Function | Perturbation type | Question it answers |
|----------|-------------------|---------------------|
| `get_network_perturbation()` | Structural (node/edge removal) | How fast does the network fall apart as nodes are lost? |
| `get_node_influence()` | Abundance-influence propagation | If I nudge one species, how far does the ripple reach? |
| `press_perturbation()` | Press (sustained-disturbance) approximation | If I keep suppressing/boosting a species, how does the community re-arrange? |

`ggnetview_perturbation_curve()` draws the attack curve produced by the
first function.

> **Interpretation caveat.** Correlation / co-occurrence networks encode
> *association, not causation*, and have no edge direction. The
> structural analysis (type 1) depends only on topology and is robust.
> The influence (type 2) and press (type 3) analyses borrow ideas from
> dynamical ecology but use correlations as proxies for interaction
> strengths -- read their output as **qualitative scenarios**, not
> quantitative predictions.

```{r setup}
library(ggNetView)
```

## 1. Build a demo network

We reuse the bundled microbiome data and build a Spearman correlation
network, exactly as in the topology vignette.

```{r build}
data(otu_rare_relative)
data(tax_tab)

mat <- as.matrix(otu_rare_relative)
mat <- mat[order(rowSums(mat), decreasing = TRUE)[seq_len(80)], ]
annot <- tax_tab[tax_tab$OTUID %in% rownames(mat), ]

g <- build_graph_from_mat(
  mat             = mat,
  method          = "cor",
  cor.method      = "spearman",
  proc            = "BH",
  r.threshold     = 0.6,
  p.threshold     = 0.05,
  module.method   = "Fast_greedy",
  node_annotation = annot,
  seed            = 1
)

g
```

## 2. Structural perturbation: random failure vs targeted attack

`get_network_perturbation()` removes nodes step by step and recomputes a
panel of connectivity metrics after each removal. The headline metric is
`LCC_fraction` -- the size of the largest connected component relative to
the original network -- and the single-number summary is the Schneider
**R-index** (area under the LCC curve; higher = more robust).

The most informative comparison is *random failure* against *targeted
attack* (removing the most-connected nodes first):

```{r structural}
rnd <- get_network_perturbation(
  g, strategy = "random", bootstrap = 30, seed = 1, plot = FALSE
)

tgt <- get_network_perturbation(
  g, strategy = "targeted", centrality = "degree", plot = FALSE
)

rbind(rnd$robustness_index, tgt$robustness_index)
```

A lower R-index under targeted attack is the classic signature of a
scale-free-like network: it tolerates random loss but is fragile to the
deliberate removal of hubs.

Overlay both curves with the companion plotting helper:

```{r structural-plot}
ggnetview_perturbation_curve(rbind(rnd$curve, tgt$curve))
```

You can target any centrality (`"degree"`, `"strength"`,
`"betweenness"`, `"closeness"`, `"eigenvector"`, or `"ivi"`), and the
`curve` table also carries `Natural_connectivity`, `Efficiency`,
`Modularity` and more, so you can plot a different metric:

```{r structural-other-metric}
ggnetview_perturbation_curve(tgt$curve, metric = "Natural_connectivity")
```

### Knocking out a specific module or node set

Instead of a fraction sweep, you can remove a named module (a level of
the `Modularity` column) or an explicit set of nodes and read the
before/after change:

```{r structural-module}
mods <- levels(get_graph_nodes(g)$Modularity)
ko <- get_network_perturbation(g, strategy = "module", target = mods[1])
ko$curve[ko$curve$metric %in% c("LCC_fraction", "N_components"), ]
```

## 3. Abundance-influence propagation

`get_node_influence()` injects a virtual perturbation at one or more
source nodes and lets it diffuse along the (signed) weighted edges,
returning a per-node `Influence` score. Positive vs negative values
distinguish neighbours that move with vs against the source.

```{r influence}
hub <- get_graph_nodes(g) |>
  dplyr::arrange(dplyr::desc(Degree)) |>
  dplyr::pull(name) |>
  head(1)

g_infl <- get_node_influence(g, source = hub)

get_graph_nodes(g_infl) |>
  dplyr::arrange(dplyr::desc(abs(Influence))) |>
  dplyr::select(name, Modularity, Degree, Influence) |>
  head(8)
```

Because the score is stored as a node column, you can map it onto a
network figure. `ggNetView()`'s `node_fill` aesthetic uses a discrete
scale, so bin the signed influence into a small categorical column
first:

```{r influence-plot}
g_infl <- g_infl |>
  tidygraph::activate(nodes) |>
  tidygraph::mutate(Influence_class = dplyr::case_when(
    Influence > 0 ~ "Co-moves (+)",
    Influence < 0 ~ "Counter-moves (-)",
    TRUE          ~ "Unreached"
  ))

ggNetView(g_infl, layout = "fr", node_fill = "Influence_class")
```

## 4. Press perturbation (sustained disturbance)

`press_perturbation()` treats the signed correlation matrix as a proxy
community matrix \(A\), adds negative self-regulation on the diagonal,
and inverts it to get the net-effect matrix \(N = -A^{-1}\). Entry
\(N_{ij}\) is the long-run net response of species \(i\) when species
\(j\) is held under sustained pressure.

The diagonal is chosen automatically to guarantee a dynamically stable
matrix, and stability is reported:

```{r press}
pp <- press_perturbation(g)

c(stable = pp$stable,
  eigen_real_max = round(pp$eigen_real_max, 3),
  self_regulation = round(pp$self_regulation, 3))
```

Query the consequences of continuously pressing the hub species -- the
nodes with the largest absolute response are the ones most affected:

```{r press-source}
pp_hub <- press_perturbation(g, source = hub)
head(pp_hub$response, 8)
```

A positive `net_response` means that species tends to increase when the
source is boosted (and decrease when it is suppressed); a negative value
means the opposite.

## Where to go next

- `vignette("network-topology")` -- static topology and Zi-Pi keystone
  roles, the natural companion to the perturbation analyses here.
- `vignette("node-importance")` -- per-node centralities and the
  integrated influence (IVI) score, which can drive the `"ivi"` targeted
  attack above.
