---
title: "Network Topology and Keystone Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Network Topology and Keystone Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi       = 72
)
```

## Overview

After building a network with any `build_graph_from_*()` function, the
natural next step is to quantify its structure. `ggNetView` provides a
family of topology functions that compute global metrics, per-sample
subgraph metrics, robustness estimates, and node-role classification:

| Function | Purpose |
|----------|---------|
| `get_network_topology()` | Global metrics + bootstrap robustness |
| `get_network_topology_parallel()` | Same, parallelised across a list of networks |
| `get_sample_subgraph_topology()` | Per-sample subgraph metrics |
| `get_sample_subgraph_topology_parallel()` | Same, parallelised |
| `ggnetview_zipi()` | Zi-Pi node-role classification |
| `ggnetview_modularity_heatmaps()` | Module composition heatmaps |

```{r setup}
library(ggNetView)
```

## 1. Build a demo network

```{r build}
data(otu_rare_relative)
data(tax_tab)

mat <- as.matrix(otu_rare_relative)
mat <- mat[order(rowSums(mat), decreasing = TRUE)[seq_len(80)], ]
annot <- tax_tab[tax_tab$OTUID %in% rownames(mat), ]

g <- build_graph_from_mat(
  mat             = mat,
  method          = "cor",
  cor.method      = "spearman",
  proc            = "BH",
  r.threshold     = 0.6,
  p.threshold     = 0.05,
  module.method   = "Fast_greedy",
  node_annotation = annot,
  seed            = 1
)

g
```

## 2. Global network topology

`get_network_topology()` returns a list with two key elements:

- **`topology`** -- a one-row data frame of global metrics (nodes,
  edges, average degree, clustering coefficient, modularity, etc.).
- **`robustness`** -- bootstrap estimates of how metrics behave under
  random node removal.

```{r global-topo}
topo <- get_network_topology(graph_obj = g, bootstrap = 30)
```

### Topology metrics

```{r topo-table}
t(topo$topology)
```

Key metrics to look for:

| Metric | Interpretation |
|--------|---------------|
| Nodes / Edges | Network size |
| Average degree | Mean connections per node |
| Clustering coefficient | Local clustering tendency |
| Modularity | Strength of community structure (> 0.4 is strong) |
| Average path length | How many hops between any two nodes |
| Density | Fraction of possible edges realised |
| Positive / Negative edges | Cooperative vs. competitive interactions |

### Robustness

The robustness table shows how topology metrics degrade as nodes are
randomly removed. This is useful for assessing network stability.

```{r robustness}
head(topo$robustness)
```

## 3. Comparing multiple networks

When you have networks from different conditions (e.g. treatment vs.
control, or different time points), pass them as a named list to
compute topology in one call.

```{r multi-topo}
mat_a <- mat[, 1:9]
mat_b <- mat[, 10:18]

g_a <- build_graph_from_mat(
  mat = mat_a, method = "cor", cor.method = "spearman",
  proc = "BH", r.threshold = 0.6, p.threshold = 0.05,
  module.method = "Fast_greedy", seed = 1
)
g_b <- build_graph_from_mat(
  mat = mat_b, method = "cor", cor.method = "spearman",
  proc = "BH", r.threshold = 0.6, p.threshold = 0.05,
  module.method = "Fast_greedy", seed = 1
)

topo_list <- get_network_topology(
  graph_obj_list = list(GroupA = g_a, GroupB = g_b),
  bootstrap      = 20
)

names(topo_list)
```

Each element contains the same `topology` + `robustness` structure.
Bind the topology rows to compare:

```{r compare-table}
topo_compare <- do.call(rbind, lapply(names(topo_list), function(nm) {
  cbind(Group = nm, topo_list[[nm]]$topology)
}))
t(topo_compare)
```

## 4. Sample-level subgraph topology

In microbiome studies, each sample contains a different subset of
taxa. `get_sample_subgraph_topology()` extracts a subgraph per sample
(keeping only taxa present in that sample) and computes topology for
each.

```{r sample-topo}
sample_topo <- get_sample_subgraph_topology(
  graph_obj = g,
  mat       = mat,
  bootstrap = 10
)

names(sample_topo)
```

The result contains:

- **`subgraph_list`** -- per-sample `tbl_graph` objects
- **`topology`** -- merged topology table (one row per sample)
- **`Robustness`** -- merged robustness table
- **`sample_stat`** -- summary of node/edge counts per sample

```{r sample-stat}
head(sample_topo$sample_stat)
```

```{r sample-topo-table}
head(sample_topo$topology)
```

### Parallel version

For large datasets with many samples, use the parallel variant to
speed up computation:

```{r sample-topo-parallel, eval = FALSE}
sample_topo_par <- get_sample_subgraph_topology_parallel(
  graph_obj = g,
  mat       = mat,
  bootstrap = 10
)
```

The output structure is identical.

## 5. Zi-Pi node-role classification

The Zi-Pi framework (Guimera & Amaral, 2005) classifies each node by
two metrics:

- **Zi (within-module connectivity)** -- how well connected a node is
  within its own module.
- **Pi (participation coefficient)** -- how much a node connects to
  other modules.

Default thresholds (Zi = 2.5, Pi = 0.62) divide nodes into four
roles:

```
                        Pi < 0.62          Pi >= 0.62
             ┌─────────────────────┬─────────────────────┐
  Zi >= 2.5  │   Module hubs       │   Network hubs      │
             ├─────────────────────┼─────────────────────┤
  Zi <  2.5  │   Peripherals       │   Connectors        │
             └─────────────────────┴─────────────────────┘
```

### Computing Zi-Pi

```{r zipi-compute}
nodes_tbl <- get_graph_nodes(g)
adj_mat   <- get_graph_adjacency(g)

zipi <- ggnetview_zipi(
  nodes_bulk     = nodes_tbl,
  z_bulk_mat     = adj_mat,
  modularity_col = "Modularity",
  degree_col     = "Degree"
)
```

The result contains a `data` frame and a ready-made `plot`:

```{r zipi-data}
head(zipi$data[, c("name", "within_module_connectivities",
                    "among_module_connectivities", "type")])
```

```{r zipi-roles}
table(zipi$data$type)
```

### Zi-Pi scatter plot

```{r zipi-plot, fig.cap = "Zi-Pi scatter plot. Quadrants identify node roles."}
zipi$plot
```

### Interpreting node roles

- **Peripherals** (most nodes) -- loosely connected within and across
  modules. They follow the community structure passively.
- **Module hubs** -- highly connected within their module but not
  bridging to others. Removal destabilises the module.
- **Connectors** -- bridge nodes linking different modules. Important
  for network-wide information flow.
- **Network hubs** -- both intra- and inter-module hubs. The most
  critical nodes; candidates for keystone species or genes.

### Custom thresholds

Adjust the classification boundaries when the defaults do not fit your
system:

```{r zipi-custom}
zipi_strict <- ggnetview_zipi(
  nodes_bulk     = nodes_tbl,
  z_bulk_mat     = adj_mat,
  modularity_col = "Modularity",
  degree_col     = "Degree",
  zi_threshold   = 2.0,
  pi_threshold   = 0.5
)

table(zipi_strict$data$type)
```

## 6. Module composition summary

You can summarise module composition by counting nodes and examining
the dominant taxa per module:

```{r mod-summary}
nodes <- get_graph_nodes(g)
module_summary <- as.data.frame(table(
  Module = as.character(nodes$Modularity)
))
colnames(module_summary)[2] <- "Nodes"
module_summary <- module_summary[order(-module_summary$Nodes), ]
module_summary
```

For richer module-environment correlation heatmaps, see
`ggnetview_modularity_heatmaps()` which links module eigengenes or
abundances to environmental variables — it requires an environmental
data frame and `env_select` specification.

## 7. Putting it all together: a typical workflow

```{r workflow-summary, eval = FALSE}
# 1. Build network
g <- build_graph_from_mat(mat, method = "cor", ...)

# 2. Visualise
ggNetView(g, layout = "fr", node_fill = "Modularity")

# 3. Global topology
topo <- get_network_topology(g, bootstrap = 100)

# 4. Sample-level topology
sample_topo <- get_sample_subgraph_topology(g, mat, bootstrap = 50)

# 5. Keystone detection
zipi <- ggnetview_zipi(
  get_graph_nodes(g), get_graph_adjacency(g),
  "Modularity", "Degree"
)

# 6. Module summary
table(get_graph_nodes(g)$Modularity)
```

Steps 3-6 work with any graph produced by `build_graph_from_mat()`,
`build_graph_from_df()`, `build_graph_from_wgcna()`, or any other
builder -- the topology functions only need a `tbl_graph` input.

## Session information

```{r sessioninfo}
sessionInfo()
```
