---
title: "WGCNA Network Pipeline"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{WGCNA Network Pipeline}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 6,
  dpi       = 72
)
has_wgcna <- requireNamespace("WGCNA", quietly = TRUE)
```

## Overview

Weighted Gene Co-expression Network Analysis (WGCNA) is the standard
approach for constructing gene co-expression networks from
high-throughput expression data. `ggNetView` integrates with WGCNA at
two levels:

1. **Quick route** -- pass `method = "WGCNA"` to `build_graph_from_mat()`
   and let ggNetView handle correlation, thresholding, and module
   detection in a single call.
2. **Full pipeline** -- run the classic WGCNA soft-thresholding and TOM
   workflow yourself, then hand the results to `trans_TOM_in_WGCNA()`
   and `build_graph_from_wgcna()` for visualisation.

Both routes produce the same `tidygraph` object that `ggNetView()` can
plot directly.

```{r setup}
library(ggNetView)
```

## 1. Quick route: `build_graph_from_mat(method = "WGCNA")`

The simplest way to build a WGCNA-style network is to switch the
`method` argument. Internally this calls `WGCNA::corAndPvalue()` for
correlation and p-values, then applies the same thresholding and module
detection as every other method.

```{r quick-route, eval = has_wgcna}
data(otu_rare_relative)
mat <- as.matrix(otu_rare_relative)
mat <- mat[order(rowSums(mat), decreasing = TRUE)[seq_len(80)], ]

graph_wgcna <- build_graph_from_mat(
  mat           = mat,
  method        = "WGCNA",
  cor.method    = "spearman",
  proc          = "BH",
  r.threshold   = 0.6,
  p.threshold   = 0.05,
  module.method = "Fast_greedy",
  seed          = 1
)

graph_wgcna
```

The returned object is identical in structure to one produced with
`method = "cor"` or `"Hmisc"`, so every downstream function
(`ggNetView()`, `get_network_topology()`, `ggnetview_zipi()`) works
unchanged.

```{r quick-plot, eval = has_wgcna, fig.cap = "Quick-route WGCNA network coloured by module."}
ggNetView(
  graph_wgcna,
  layout    = "fr",
  seed      = 1,
  node_size_range = c(2, 7),
  node_fill   = "Modularity",
  module_label     = FALSE
)
```

### Attaching taxonomy

Pass a node annotation table to colour by taxonomy instead of module:

```{r quick-annot, eval = has_wgcna}
data(tax_tab)
annot <- tax_tab[tax_tab$OTUID %in% rownames(mat), ]

graph_wgcna_annot <- build_graph_from_mat(
  mat             = mat,
  method          = "WGCNA",
  cor.method      = "spearman",
  proc            = "BH",
  r.threshold     = 0.6,
  p.threshold     = 0.05,
  module.method   = "Fast_greedy",
  node_annotation = annot,
  seed            = 1
)
```

```{r quick-annot-plot, eval = has_wgcna, fig.cap = "WGCNA network coloured by Phylum."}
ggNetView(
  graph_wgcna_annot,
  layout    = "fr",
  seed      = 1,
  node_size_range = c(2, 7),
  node_fill   = "Phylum",
  module_label     = FALSE
)
```

## 2. Full WGCNA pipeline

When you need fine-grained control over soft-thresholding power, TOM
type, or the dendrogram cut height, run the classic WGCNA steps
yourself and feed the results into ggNetView.

### Step 1: Prepare expression data

WGCNA expects **samples in rows** and **genes/features in columns**.

```{r full-prepare, eval = has_wgcna}
expr_mat <- t(mat)
dim(expr_mat)
```

### Step 2: Pick a soft-thresholding power

```{r full-power, eval = has_wgcna, fig.cap = "Scale-free topology fit versus soft-thresholding power."}
# Force WGCNA into single-threaded mode before calling
# `pickSoftThreshold()`. WGCNA spins up an internal foreach/parallel
# cluster by default, which is fragile inside vignette / R CMD check
# environments (and intermittently dies with
# "Error in summary.connection(connection) : invalid connection").
# Single-threaded is slower but deterministic and side-effect-free,
# which is exactly what we want for reproducible documentation. Users
# can re-enable parallelism for production work via
# `WGCNA::allowWGCNAThreads()` or `WGCNA::enableWGCNAThreads()`.
WGCNA::disableWGCNAThreads()

powers <- c(seq(1, 10, by = 1), seq(12, 30, by = 2))
sft <- WGCNA::pickSoftThreshold(
  expr_mat,
  powerVector  = powers,
  networkType  = "signed",
  verbose      = 0
)

plot(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     xlab = "Soft Threshold (power)",
     ylab = "Scale Free Topology Model Fit (signed R^2)",
     main = "Scale independence",
     type = "n")
text(sft$fitIndices[, 1],
     -sign(sft$fitIndices[, 3]) * sft$fitIndices[, 2],
     labels = powers, col = "red")
abline(h = 0.85, col = "red")
```

Choose the lowest power where the model fit ($R^2$) exceeds 0.85.

```{r full-pick-power, eval = has_wgcna}
picked_power <- sft$powerEstimate
if (is.na(picked_power)) picked_power <- 6
message("Using soft-thresholding power: ", picked_power)
```

### Step 3: Build TOM and detect modules

```{r full-tom, eval = has_wgcna}
adjacency <- WGCNA::adjacency(
  expr_mat,
  power       = picked_power,
  type        = "signed",
  corFnc      = "cor",
  corOptions  = list(method = "spearman", use = "pairwise.complete.obs")
)

TOM <- WGCNA::TOMsimilarity(adjacency, TOMType = "signed")
dissTOM <- 1 - TOM

gene_tree <- hclust(as.dist(dissTOM), method = "average")

dynamic_mods <- dynamicTreeCut::cutreeDynamic(
  dendro    = gene_tree,
  distM     = dissTOM,
  deepSplit = 2,
  minClusterSize = 5
)

module_colors <- WGCNA::labels2colors(dynamic_mods)
module_df <- data.frame(
  ID     = colnames(expr_mat),
  Module = module_colors,
  stringsAsFactors = FALSE
)
head(module_df)
```

### Step 4: Convert TOM to edge list

`trans_TOM_in_WGCNA()` converts the dense TOM matrix into a long-format
edge list (`from`, `to`, `weight`). Use the `threshold` argument to
keep only the strongest edges:

```{r full-trans, eval = has_wgcna}
# `trans_TOM_in_WGCNA()` uses `colnames(mat)` as node IDs, so `mat` must
# be in the same orientation as the input to `WGCNA::adjacency()`:
# samples in rows, features (genes/OTUs) in columns. `expr_mat` is
# already in that shape (we transposed once at Step 1), so pass it
# straight in -- transposing again would swap node IDs to sample IDs and
# fail the dimension check inside trans_TOM_in_WGCNA().
edge_df <- trans_TOM_in_WGCNA(
  TOM       = TOM,
  mat       = expr_mat,
  threshold = 0.1
)
head(edge_df)
dim(edge_df)
```

### Step 5: Build the ggNetView graph object

```{r full-build, eval = has_wgcna}
graph_full <- build_graph_from_wgcna(
  wgcna_tom       = edge_df,
  module          = module_df,
  node_annotation = annot,
  seed            = 1
)
graph_full
```

### Step 6: Visualise

```{r full-plot-fr, eval = has_wgcna, fig.cap = "Full WGCNA pipeline -- Fruchterman-Reingold layout."}
ggNetView(
  graph_full,
  layout    = "fr",
  seed      = 1,
  node_size_range = c(2, 7),
  node_fill   = "Modularity",
  module_label     = FALSE
)
```

Try different layouts to highlight module structure:

```{r full-plot-circular, eval = has_wgcna, fig.cap = "Circular module layout groups nodes by WGCNA module."}
# Note the `_layout` suffix: the petal-style layout functions are named
# `create_layout_circular_modules_petal_layout()` (and the *2 variant),
# whereas other layouts like `fr` / `gephi` / `kk` use the bare name. The
# `layout` argument is concatenated to `create_layout_` to look up the
# function, so it must include the suffix here.
ggNetView(
  graph_full,
  layout    = "circular_modules_petal_layout",
  seed      = 1,
  node_size_range = c(2, 6),
  node_fill   = "Modularity",
  module_label     = FALSE
)
```

## 3. RMT-guided threshold selection

When you are unsure what `r.threshold` to use, Random Matrix Theory
(RMT) can pick one automatically. `ggNetView_RMT()` scans a range of
thresholds and selects the one whose nearest-neighbour spacing
distribution best fits the Poisson (non-random) model.

```{r rmt-scan, eval = has_wgcna, fig.cap = "RMT auto-score across thresholds."}
rmt_res <- ggNetView_RMT(
  mat     = mat,
  method  = "WGCNA",
  cor.method = "spearman",
  nr_thresholds = 31,
  verbose = FALSE,
  seed    = 1
)

message("RMT chosen threshold: ", round(rmt_res$chosen_threshold, 4))
```

Feed the chosen threshold back into the builder:

```{r rmt-build, eval = has_wgcna}
graph_rmt <- build_graph_from_mat(
  mat           = mat,
  method        = "WGCNA",
  cor.method    = "spearman",
  proc          = "BH",
  r.threshold   = rmt_res$chosen_threshold,
  p.threshold   = 0.05,
  module.method = "Fast_greedy",
  seed          = 1
)
```

```{r rmt-plot, eval = has_wgcna, fig.cap = "Network built with RMT-selected threshold."}
ggNetView(
  graph_rmt,
  layout    = "fr",
  seed      = 1,
  node_size_range = c(2, 7),
  node_fill   = "Modularity",
  module_label     = FALSE
)
```

### Visualising the RMT scan

The `scores` element contains per-threshold diagnostics. A quick plot
helps verify the choice:

```{r rmt-scores-plot, eval = has_wgcna, fig.cap = "KS p-value (higher = more Poisson-like) across thresholds."}
scores <- rmt_res$scores
plot(scores$threshold, scores$ks_p,
     type = "b", pch = 16,
     xlab = "Correlation threshold",
     ylab = "KS p-value (Exponential fit)",
     main = "RMT threshold scan")
abline(v = rmt_res$chosen_threshold, lty = 2, col = "red")
```

## 4. Topology and keystone analysis

Once you have a graph object (from either route), the topology and
Zi-Pi functions work identically.

### Global topology

```{r topology, eval = has_wgcna}
topo <- get_network_topology(graph_obj = graph_wgcna, bootstrap = 20)
head(topo$topology)
```

### Zi-Pi classification

```{r zipi, eval = has_wgcna}
nodes_tbl <- get_graph_nodes(graph_wgcna)
adj_mat   <- get_graph_adjacency(graph_wgcna)

zipi <- ggnetview_zipi(
  nodes_bulk     = nodes_tbl,
  z_bulk_mat     = adj_mat,
  modularity_col = "Modularity",
  degree_col     = "Degree"
)
```

```{r zipi-plot, eval = has_wgcna, fig.cap = "Zi-Pi scatter plot identifying keystone taxa."}
zipi$plot
```

Nodes in the upper-right quadrant (high Zi **and** high Pi) are
**network hubs** -- candidates for keystone species or genes that bridge
multiple functional modules.

## 5. Quick-route vs full-pipeline: when to use which

| Feature | Quick route | Full pipeline |
|---------|-------------|---------------|
| Lines of code | ~10 | ~40 |
| Soft-thresholding power | Not applicable (hard threshold) | User-selected |
| TOM-based edges | No (correlation-based) | Yes |
| Module detection | igraph algorithms | WGCNA `dynamicTreeCut` |
| Best for | Exploratory analysis, microbiome data | Gene expression, publication-grade networks |

The **quick route** treats WGCNA purely as a fast correlation engine and
applies the same hard-threshold + igraph-module pipeline as all other
methods. The **full pipeline** preserves the topological overlap
information and WGCNA's own hierarchical module detection, which is
the standard for gene co-expression studies.

## Session information

```{r sessioninfo}
sessionInfo()
```
